Synthesising Recursive Functions for First-Order Model Counting

### 2023-02-06 doc/paper/paper.tex
3/4 proofs complete

% See https://www.overleaf.com/learn/latex/theorems_and_proofs
% for a nice explanation of how to define new theorems, but keep
% in mind that the amsthm package is already included in this
% template and that you must *not* alter the styling.
% \newtheorem{example}{Example}
% \newtheorem{theorem}{Theorem}
\newtheorem{theorem}{Theorem}
\newtheorem{conjecture}{Conjecture}
\newtheorem{claim}{Claim}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}

@@

% TODO: read the formatting instructions and their LaTeX source (email font?)
% TODO: make sure there are no overflow errors
% TODO: maybe mention the term 'data complexity'
% TODO: make claim numbers subordinate to theorem numbers
% TODO: mention that most of the work for this paper was done at UoE
% TODO: the appendix should be in a separate file
% TODO: cite some works about synthesis
% TODO: clean up unnecessary things from the header (e.g., packages and definitions)
% TODO: don't use the '\in' syntax for predicates
% TODO: make sure I'm using \sigma instead of || everywhere
% TODO: (at the end): grammarly the whole thing, put the appendix in a separate file
\author{%
Paulius Dilkas$^1$\and
Vaishak Belle$^2$\\
\affiliations
$^1$National University of Singapore, Singapore, Singapore\\
$^2$University of Edinburgh, Edinburgh, UK\\
\emails
paulius.dilkas@nus.edu.sg,
vbelle@ed.ac.uk
}

@@

@@

One can read a formula in our format as a sentence in FOL\@. All variables in a
clause are implicitly universally quantified (but note that variables are never
shared among clauses), and all clauses in a formula are implicitly linked by a
conjunction. Thus, formula $\phi$ from \cref{example:first} reads as
\begin{align*}
  &\begin{multlined}[t]
    (\forall X \in \Gamma\text{. }\forall Y, Z \in \Delta\text{. }\\
    Y \ne Z \implies \neg p(X, Y) \lor \neg p(X, Z)) \land
    \end{multlined}\\
  &\begin{multlined}[t]
    (\forall X, Z \in \Gamma\text{. }\forall Y \in \Delta\text{. }\\
    X \ne Z \implies \neg p(X, Y) \lor \neg p(Z, Y)).
    \end{multlined}
\end{align*}

@@

\begin{algorithm}[t]
  \caption{The compilation rule for $\GDR$ nodes.}\label{alg:domainrecursion}
  \KwIn{formula $\phi$, set of all relevant domains $\mathcal{D}$}
  \KwOut{set of chips $S$}
  $S \gets \emptyset$\;
  \ForEach{domain $\Omega \in \mathcal{D}$ s.t.\ there is $c \in \phi$ and $X \in \Vars(L_c)$ s.t. $\delta_c(X) = \Omega$\label{line:condition}}{
    $\phi' \gets \emptyset$\;
    $x \gets \text{a new constant in domain } \Omega$\;
    \ForEach{clause $c = (L, C, \delta) \in \phi$\label{line:forclause}}{
      $V \gets \{\, X \in \Vars(L) \mid \delta(X) = \Omega \,\}$\;
      \ForEach{subset $W \subseteq V$ s.t. $W^2 \cap C = \emptyset$ {\bf and} $W \cap \{\, X \in \Vars(C) \mid (X, y) \in C \text{ for some constant } y \,\} = \emptyset$\label{line:conditions}}{
        \tcc{$\delta'$ restricts $\delta$ to the new set of variables}
        $\phi' \gets \phi' \cup \{\, (L[x/W], C[x/W] \cup \{\, (X, x) \mid (X \in V \setminus W) \,\}, \delta') \,\}$\;\label{line:generation}
      }
    }
    $S \gets S \cup \{\, (\GDR_{\Omega \gets \Omega \setminus \{\, x \,\}}, \langle\phi'\rangle) \,\}$\;
  }
\end{algorithm}

@@

\begin{theorem}[Correctness of $\GDR$]
  Let $\phi$ be the formula used as input to \cref{alg:domainrecursion} and
  $\phi'$ be one of the formulas constructed by the algorithm. Then
  $\phi \equiv \phi'$.
\end{theorem}
\begin{proof} % TODO: finish (Proof technique: if one is true then the other one will be true and vice versa)
  Let $\Omega \in \mathcal{D}$ be the domain selected by
  \cref{line:condition} of the algorithm.
  \begin{claim}\label{claim:logic}
  \[
    \forall X \in \Omega\text{. } \psi \equiv \psi[x/\{\, X \,\}] \land \forall X \in \Omega\text{. } X \ne x \implies \psi
  \]
  \end{claim}
  \begin{claim}
    Iteration of \cref{claim:logic} for all variables in some set of variables
    $V$ produces the same clauses as subset enumeration on
    \cref{line:conditions}.
  \end{claim}
\end{proof}

@@

  If $c$ has no variables with domain $\Omega$, then it cannot have any
  constraints involving $x$, so $c' = c$. Otherwise, for notational simplicity,
  let us assume that $X$ is the only variable in $c$ with domain $\Omega$ (the
  proof for an arbitrary number of variables is virtually the same). By
  \cref{def:replaceable}, we can rewrite $c$ as $\forall X \in \Omega\text{.
  } X \ne x \implies \psi$, where $\psi$ is a formula with $X$ as the only free
  variable and with no mention of either $x$ or $\Omega$. Then
  $c' \equiv \forall X \in \Omega'\text{. } \psi$. Since
  $\Omega' \coloneqq \Omega \setminus \{\, x \,\}$, we have that $c \equiv c'$.
  Since $c$ was an arbitrary clause of $\phi$, this completes the proof that
  $\phi \equiv \phi'$.
\end{proof}

@@

\section{How to Interpret an FCG}\label{sec:interpret}

@@

  A contradiction node with clause $c$ as a parameter is interpreted as one if
  the clause has groundings and zero otherwise. In this case, the parameter is
  $c = (\emptyset, \{\, (X, Y) \,\}, \{\, X \mapsto \Delta^\top, Y \mapsto \Delta^\top \,\})$
  (not shown in \cref{fig:examplefcg}), which can be read as
  $\forall X, Y \in \Delta^{\top}\text{. }X \ne Y \implies \bot$, i.e.,
  $\forall X, Y \in \Delta^{\top}\text{. }X = Y$. This latter sentence is true
  if and only if $|\Delta^{\top}| < 2$. Therefore, we can use the Iverson
  bracket notation to write
  \[
    \diamondsuit = [l < 2] \coloneqq
    \begin{cases}
      1 & \text{if } l < 2 \\
      0 & \text{otherwise.}
    \end{cases}
  \]

@@

% Describe what kind of instances we're investigating
Let $p \in \Gamma \times \Delta$ be a predicate. To restrict all relations
representable by $p$ to just functions from $\Gamma$ to $\Delta$, in FOL one
might write
\[
  \forall X \in \Gamma\text{. }\forall Y,Z \in \Delta\text{. }p(X, Y) \land p(X, Z) \implies Y = Z
\]
and
\begin{equation}\label{eq:def2}
  \forall X \in \Gamma\text{. }\exists Y \in \Delta\text{. }p(X, Y).
\end{equation}
The former sentence says that one element of $\Gamma$ can map to at \emph{most}
one element of $\Delta$, and the latter sentence says that each element of
$\Gamma$ must map to at \emph{least} one element of $\Delta$. One can then add
\[
  \forall X,Z \in \Gamma\text{. }\forall Y \in \Delta\text{. }p(X, Y) \land p(Z, Y) \implies X = Z
\]
to restrict $p$ to injections or
\[
  \forall Y \in \Delta\text{. }\exists X \in \Gamma\text{. }p(X, Y)
\]
to ensure surjectivity or remove \cref{eq:def2} to consider partial functions.
Lastly, one can replace all occurrences of $\Delta$ with $\Gamma$ to model
endofunctions (i.e., functions with the same domain and codomain) instead.

@@

\begin{conjecture}
  Let \IFO{} be the class of formulas in FOL that contain clauses with at most
  two variables as well as any number of copies of
  \begin{align*}
    &\begin{multlined}[t]
      (\forall X \in \Gamma\text{. }\forall Y, Z \in \Delta\text{. }\\
      Y \ne Z \implies \neg p(X, Y) \lor \neg p(X, Z)) \land
    \end{multlined}\\
    &\begin{multlined}[t]
      (\forall X, Z \in \Gamma\text{. }\forall Y \in \Delta\text{. }\\
      X \ne Z \implies \neg p(X, Y) \lor \neg p(Z, Y))
    \end{multlined}
  \end{align*}
  for some predicate $p$ and domains $\Gamma$ and $\Delta$. Then \IFO{} is
  liftable by \textsc{Crane}.
\end{conjecture}

>>>>

% See https://www.overleaf.com/learn/latex/theorems_and_proofs
% for a nice explanation of how to define new theorems, but keep
% in mind that the amsthm package is already included in this
% template and that you must *not* alter the styling.
% \newtheorem{example}{Example}
% \newtheorem{theorem}{Theorem}
\newtheorem{theorem}{Theorem}
\newtheorem{conjecture}{Conjecture}
\newtheorem{claim}{Claim}[theorem]
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}

@@

% TODO (at the very end)
% put the appendix in a separate file
% grammarly the whole thing
% read the formatting instructions and their LaTeX source (email font?)
% make sure there are no overflow errors
% clean up unnecessary things from the header (e.g., packages and definitions)

\author{%
Paulius Dilkas$^1$\and
Vaishak Belle$^2$\\
\affiliations
$^1$National University of Singapore, Singapore, Singapore\\
$^2$University of Edinburgh, Edinburgh, UK\\
\emails
paulius.dilkas@nus.edu.sg,
vbelle@ed.ac.uk
}

@@

% TODO: maybe mention the term 'data complexity'
% TODO: cite some works about synthesis

@@

One can read a formula in our format as a sentence in FOL\@. All variables in a
clause are implicitly universally quantified (but note that variables are never
shared among clauses), and all clauses in a formula are implicitly linked by a
conjunction. Thus, formula $\phi$ from \cref{example:first} reads as
\begin{align*}
  &\begin{multlined}[t]
    (\forall X \in \Gamma\text{. }\forall Y, Z \in \Delta\text{. }\\
    Y \ne Z \Rightarrow \neg p(X, Y) \lor \neg p(X, Z)) \land
    \end{multlined}\\
  &\begin{multlined}[t]
    (\forall X, Z \in \Gamma\text{. }\forall Y \in \Delta\text{. }\\
    X \ne Z \Rightarrow \neg p(X, Y) \lor \neg p(Z, Y)).
    \end{multlined}
\end{align*}

@@

\begin{algorithm}[t]
  \caption{The compilation rule for $\GDR$ nodes.}\label{alg:domainrecursion}
  \KwIn{formula $\phi$, set of all relevant domains $\mathcal{D}$}
  \KwOut{set of chips $S$}
  $S \gets \emptyset$\;
  \ForEach{domain $\Omega \in \mathcal{D}$ s.t.\ there is $c \in \phi$ and $X \in \Vars(L_c)$ s.t. $\delta_c(X) = \Omega$\label{line:condition}}{
    $\phi' \gets \emptyset$\;
    $x \gets \text{a new constant in domain } \Omega$\;\label{line:constant}
    \ForEach{clause $c = (L, C, \delta) \in \phi$\label{line:forclause}}{
      $V \gets \{\, X \in \Vars(L) \mid \delta(X) = \Omega \,\}$\;\label{line:V}
      \ForEach{$W \subseteq V$ s.t. $W^2 \cap C = \emptyset$ {\bf and} $W \cap \{\, X \in \Vars(C) \mid (X, y) \in C \text{ for some constant } y \,\} = \emptyset$\label{line:conditions}}{
        \tcc{$\delta'$ restricts $\delta$ to the new set of variables}
        $\phi' \gets \phi' \cup \{\, (L[x/W], C[x/W] \cup \{\, (X, x) \mid (X \in V \setminus W) \,\}, \delta') \,\}$\;\label{line:generation}
      }
    }
    $S \gets S \cup \{\, (\GDR_{\Omega \gets \Omega \setminus \{\, x \,\}}, \langle\phi'\rangle) \,\}$\;
  }
\end{algorithm}

@@

\begin{theorem}[Correctness of $\GDR$\footnote{\Cref{thm:correctness1} is equivalent to
    Proposition~3 by~\cite{DBLP:conf/nips/Broeck11}.}]\label{thm:correctness1}
  Let $\phi$ be the formula used as input to \cref{alg:domainrecursion},
  $\Omega \in \mathcal{D}$ the domain selected on \cref{line:condition}, and
  $\phi'$ the formula constructed by the algorithm for $\Omega$. Assume that
  $\Omega \ne \emptyset$. Then $\phi \equiv \phi'$.
\end{theorem}
\begin{proof}
  Let $x$ be the constant introduced on \cref{line:constant} and $c$ a clause of
  $\phi$ selected on \cref{line:forclause} of the algorithm. We will show that
  $c$ is equivalent to the conjunction of clauses added to $\phi'$ on
  \cref{line:conditions,line:generation}.

  If $c$ has no variables with domain $\Omega$, then
  \cref{line:conditions,line:generation} simply add a copy of $c$ to $\phi'$. If
  $c$ has one variable with domain $\Omega$, say, $X$, then
  $c \equiv \forall X \in \Omega\text{. } \psi$ for some formula $\psi$ with one
  free variable $X$. In this case, \cref{line:conditions,line:generation} create
  two clauses: $\psi[x/\{\, X \,\}]$ and $\forall X \in \Omega\text{.
  } X \ne x \Rightarrow \psi$. It is easy to see that their conjunction is
  equivalent to $c$ provided that $\Omega \ne \emptyset$ and so $x$ is a
  well-defined constant.

  Let $V$ be the set of variables introduced on \cref{line:V}. It remains to
  show that repeating the transformation
  \[
    \forall X \in \Omega\text{. } \psi \mapsto \{\, \psi[x/\{\, X \,\}], \forall X \in \Omega\text{. } X \ne x \Rightarrow \psi \,\}
  \]
  for all variables $X \in V$ produces the same clauses as
  \cref{line:conditions,line:generation} (except for some tautologies that the
  latter method skips). Note that the order in which these operations are
  applied is immaterial. In other words, if we add inequality constraints for
  variables ${\{\, X_{i} \,\}}_{i=1}^{n}$, and apply substitution for variables
  $V \setminus {\{\, X_{i} \,\}}_{i=1}^{n}$, then---regardless of the order of
  operations---we get
  \[
    \forall X_{1}, \dots, X_{n} \in \Omega\text{.} \bigwedge_{i=1}^{n} X_{i} \ne x \Rightarrow \psi\left[x / \left(V \setminus {\{\, X_{i} \,\}}_{i=1}^{n}\right)\right].
  \]
  This formula is equivalent to the clause generated on \cref{line:generation}
  with $W = V \setminus {\{\, X_{i} \,\}}_{i=1}^{n}$. Thus, for every clause $c$
  of $\phi$, the new formula $\phi'$ gets a set of clauses whose conjunction is
  equivalent to $c$. Hence, the two formulas are equivalent.
\end{proof}

@@

  If $c$ has no variables with domain $\Omega$, then it cannot have any
  constraints involving $x$, so $c' = c$. Otherwise, for notational simplicity,
  let us assume that $X$ is the only variable in $c$ with domain $\Omega$ (the
  proof for an arbitrary number of variables is virtually the same). By
  \cref{def:replaceable}, we can rewrite $c$ as $\forall X \in \Omega\text{.
  } X \ne x \Rightarrow \psi$, where $\psi$ is a formula with $X$ as the only free
  variable and with no mention of either $x$ or $\Omega$. Then
  $c' \equiv \forall X \in \Omega'\text{. } \psi$. Since
  $\Omega' \coloneqq \Omega \setminus \{\, x \,\}$, we have that $c \equiv c'$.
  Since $c$ was an arbitrary clause of $\phi$, this completes the proof that
  $\phi \equiv \phi'$.
\end{proof}

@@

\section{How to Interpret an FCG}\label{sec:interpret}
% TODO: make sure I'm using \sigma instead of || in this section

@@

  A contradiction node with clause $c$ as a parameter is interpreted as one if
  the clause has groundings and zero otherwise. In this case, the parameter is
  $c = (\emptyset, \{\, (X, Y) \,\}, \{\, X \mapsto \Delta^\top, Y \mapsto \Delta^\top \,\})$
  (not shown in \cref{fig:examplefcg}), which can be read as
  $\forall X, Y \in \Delta^{\top}\text{. }X \ne Y \Rightarrow \bot$, i.e.,
  $\forall X, Y \in \Delta^{\top}\text{. }X = Y$. This latter sentence is true
  if and only if $|\Delta^{\top}| < 2$. Therefore, we can use the Iverson
  bracket notation to write
  \[
    \diamondsuit = [l < 2] \coloneqq
    \begin{cases}
      1 & \text{if } l < 2 \\
      0 & \text{otherwise.}
    \end{cases}
  \]

@@

% TODO: don't use the \in syntax
% Describe what kind of instances we're investigating
Let $p \in \Gamma \times \Delta$ be a predicate. To restrict all relations
representable by $p$ to just functions from $\Gamma$ to $\Delta$, in FOL one
might write
\[
  \forall X \in \Gamma\text{. }\forall Y,Z \in \Delta\text{. }p(X, Y) \land p(X, Z) \Rightarrow Y = Z
\]
and
\begin{equation}\label{eq:def2}
  \forall X \in \Gamma\text{. }\exists Y \in \Delta\text{. }p(X, Y).
\end{equation}
The former sentence says that one element of $\Gamma$ can map to at \emph{most}
one element of $\Delta$, and the latter sentence says that each element of
$\Gamma$ must map to at \emph{least} one element of $\Delta$. One can then add
\[
  \forall X,Z \in \Gamma\text{. }\forall Y \in \Delta\text{. }p(X, Y) \land p(Z, Y) \Rightarrow X = Z
\]
to restrict $p$ to injections or
\[
  \forall Y \in \Delta\text{. }\exists X \in \Gamma\text{. }p(X, Y)
\]
to ensure surjectivity or remove \cref{eq:def2} to consider partial functions.
Lastly, one can replace all occurrences of $\Delta$ with $\Gamma$ to model
endofunctions (i.e., functions with the same domain and codomain) instead.

@@

\begin{conjecture}
  Let \IFO{} be the class of formulas in FOL that contain clauses with at most
  two variables as well as any number of copies of
  \begin{align*}
    &\begin{multlined}[t]
      (\forall X \in \Gamma\text{. }\forall Y, Z \in \Delta\text{. }\\
      Y \ne Z \Rightarrow \neg p(X, Y) \lor \neg p(X, Z)) \land
    \end{multlined}\\
    &\begin{multlined}[t]
      (\forall X, Z \in \Gamma\text{. }\forall Y \in \Delta\text{. }\\
      X \ne Z \Rightarrow \neg p(X, Y) \lor \neg p(Z, Y))
    \end{multlined}
  \end{align*}
  for some predicate $p$ and domains $\Gamma$ and $\Delta$. Then \IFO{} is
  liftable by \textsc{Crane}.
\end{conjecture}
